---
title: "Final"
author: "Lea Lambdin"
format: pdf
GitHub Repository:

https://github.com/LeaLambdin/ENVS-193DS_spring-2026_final
---

```{r}
```{r}
library(tidyverse)
library(janitor)
library(rstatix)
library(ggpubr)

# read data here once downloaded
# example:
# data <- read_csv("../data/datafile.csv")
```

## Problem 1

### a.

In part 1, they likely used a logistic regression. The response variable is whether American Avocets used a microhabitat (yes/no), and the predictor variable is distance to the water edge. Logistic regression is commonly used when the response variable is binary and the goal is to determine whether a predictor affects the probability of an outcome.

In part 2, they likely used a chi-square test of independence. The variables are bird species (categorical) and habitat type (categorical), and the test evaluates whether there is an association between the two categorical variables.

### b.

Part 1 rewrite: 

American Avocet microhabitat use changed with distance from the water edge. Areas closer to or farther from the water edge were used differently by American Avocets.

Part 2 rewrite: 

Habitat use differed among the three bird species. American Avocets, Forster’s Terns, and Black-necked Stilts were associated with different habitat types within the San Francisco Bay.

### c.

One additional variable that could influence American Avocet microhabitat use is native plant percent cover, measured as the percent of ground covered by native plants within a sampling plot. This is a continuous numeric variable measured in percent, and it could influence habitat use because plant cover may affect nesting protection, visibility, predator risk, and access to foraging areas.
