---
title: "Final"
author: "Lea Lambdin"
format: pdf
GitHub Repository:

https://github.com/LeaLambdin/ENVS-193DS_spring-2026_final
---

```{r}

#install packages
library(tidyverse)
library(janitor)
library(rstatix)

#read in data
nests <- read_csv("data/nest_data_final.csv")
```


## Problem 1

### a.

In part 1, they likely used a logistic regression. The response variable is whether American Avocets used a microhabitat (yes/no), and the predictor variable is distance to the water edge. Logistic regression is commonly used when the response variable is binary and the goal is to determine whether a predictor affects the probability of an outcome.

In part 2, they likely used a chi-square test of independence. The variables are bird species (categorical) and habitat type (categorical), and the test evaluates whether there is an association between the two categorical variables.

### b.

Part 1 rewrite: 

American Avocet microhabitat use changed with distance from the water edge. Areas closer to or farther from the water edge were used differently by American Avocets.

Part 2 rewrite: 

Habitat use differed among the three bird species. American Avocets, Forster’s Terns, and Black-necked Stilts were associated with different habitat types within the San Francisco Bay.

### c.

One additional variable that could influence American Avocet microhabitat use is native plant percent cover, measured as the percent of ground covered by native plants within a sampling plot. This is a continuous numeric variable measured in percent, and it could influence habitat use because plant cover may affect nesting protection, visibility, predator risk, and access to foraging areas.

## Problem 2

### a.

```{r}
nests_clean <- nests |>

  # keep variables needed for analysis
  select(case_control, height_cm, ant_species) |>

  # keep only the three focal ant species
  filter(ant_species %in% c("AB", "RRB", "BBR")) |>

  # reorder factor levels and add full species names
  mutate(
    ant_species = factor(
      ant_species,
      levels = c("AB", "RRB", "BBR")
    ),
    ant_species_full = case_when(
      ant_species == "AB" ~ "Crematogaster sjostedti",
      ant_species == "RRB" ~ "Crematogaster mimosae",
      ant_species == "BBR" ~ "Crematogaster nigriceps"
    )
  )

# display structure of cleaned data
str(nests_clean)

# display 10 random rows
slice_sample(nests_clean, n = 10)
```

### b.

In this data set, a value of 1 indicates that a bird nest was present in the tree, while a value of 0 indicates that no bird nest was present. Biologically, these values represent whether a tree was used or not used as nesting habitat by birds.


