---
title: "Final"
author: "Lea Lambdin"
format: pdf
GitHub Repository:

https://github.com/LeaLambdin/ENVS-193DS_spring-2026_final
---

```{r}

#install packages
library(tidyverse)
library(janitor)
library(rstatix)
library(DHARMa)

#read in data
nests <- read_csv("data/nest_data_final.csv")
```


## Problem 1

### a.

In part 1, they likely used a logistic regression. The response variable is whether American Avocets used a microhabitat (yes/no), and the predictor variable is distance to the water edge. Logistic regression is commonly used when the response variable is binary and the goal is to determine whether a predictor affects the probability of an outcome.

In part 2, they likely used a chi-square test of independence. The variables are bird species (categorical) and habitat type (categorical), and the test evaluates whether there is an association between the two categorical variables.

### b.

Part 1 rewrite: 

American Avocet microhabitat use changed with distance from the water edge. Areas closer to or farther from the water edge were used differently by American Avocets.

Part 2 rewrite: 

Habitat use differed among the three bird species. American Avocets, Forster’s Terns, and Black-necked Stilts were associated with different habitat types within the San Francisco Bay.

### c.

One additional variable that could influence American Avocet microhabitat use is native plant percent cover, measured as the percent of ground covered by native plants within a sampling plot. This is a continuous numeric variable measured in percent, and it could influence habitat use because plant cover may affect nesting protection, visibility, predator risk, and access to foraging areas.

## Problem 2

### a.

```{r}
nests_clean <- nests |>

  # keep variables needed for analysis
  select(case_control, height_cm, ant_species) |>

  # keep only the three focal ant species
  filter(ant_species %in% c("AB", "RRB", "BBR")) |>

  # reorder factor levels and add full species names
  mutate(
    ant_species = factor(
      ant_species,
      levels = c("AB", "RRB", "BBR")
    ),
    ant_species_full = case_when(
      ant_species == "AB" ~ "Crematogaster sjostedti",
      ant_species == "RRB" ~ "Crematogaster mimosae",
      ant_species == "BBR" ~ "Crematogaster nigriceps"
    )
  )

# display structure of cleaned data
str(nests_clean)

# display 10 random rows
slice_sample(nests_clean, n = 10)
```

### b.

In this data set, a value of 1 indicates that a bird nest was present in the tree, while a value of 0 indicates that no bird nest was present. Biologically, these values represent whether a tree was used or not used as nesting habitat by birds.

### c. 

Tree height

Tree height is a continuous variable measured in centimeters. As tree height increases, the probability of nest occurrence may increase because taller trees provide more space and structure for nesting and may offer better protection from predators. This information was found in the Methods section (nest tree surveys/data collection) where tree characteristics, including height, were measured for nest and available trees.

Acacia ant species

Acacia ant species is a categorical variable with three levels in this analysis: Crematogaster sjostedti, Crematogaster mimosae, and Crematogaster nigriceps. Ant species may influence the probability of nest occurrence because different ant species vary in how aggressively they defend trees from browsing mammals, and birds preferentially nested in trees occupied by strongly defensive ant species. This information was discussed in the Introduction and study rationale, where the authors explain the relationship between ant mutualists and bird nest-site selection.

### d.

```{r}
# create a table describing the models that will be compared
model_table <- tibble(

  # model identifier
  model_number = c(1, 2),

  # response variable for each model
  response_variable = c(
    "case_control",
    "case_control"
  ),

  # predictors included in each model
  predictors = c(
    "height_cm + ant_species",
    "height_cm + ant_species"
  ),

  # indicate whether an interaction term is included
  interaction_term = c(
    "No",
    "height_cm × ant_species"
  )
)

# display model table
model_table
```

### e.

```{r}
# run logistic regression model without interaction
# this model tests whether tree height and ant species predict nest occurrence
model_1 <- glm(
  case_control ~ height_cm + ant_species,
  data = nests_clean,
  family = binomial
)

# run logistic regression model with interaction
# this model tests whether the effect of tree height depends on ant species
model_2 <- glm(
  case_control ~ height_cm * ant_species,
  data = nests_clean,
  family = binomial
)
```

### f.

```{r}
# compare AIC values for the two candidate models
AIC(model_1, model_2)
```

The best model as determined by Akaike's Information Criterion (AIC) included tree height, acacia ant species, and the interaction between tree height and acacia ant species as predictors of nest occurrence. This model had a lower AIC value than the model that included only the main effects of tree height and acacia ant species.

### g. 

```{r, fig.width=7, fig.height=6}
# simulate residuals for the selected logistic regression model
# the selected model includes tree height, ant species, and their interaction
model_2_residuals <- simulateResiduals(
  fittedModel = model_2
)

# display diagnostic plots for the simulated residuals
plot(model_2_residuals)
```

### h.

```{r}
# create prediction data using the same ant species levels as the model
prediction_data <- expand_grid(
  height_cm = seq(
    min(nests_clean$height_cm),
    max(nests_clean$height_cm),
    length.out = 100
  ),
  ant_species = levels(nests_clean$ant_species)
) |>
  mutate(
    ant_species = factor(
      ant_species,
      levels = levels(nests_clean$ant_species)
    ),
    ant_species_full = case_when(
      ant_species == "AB" ~ "Crematogaster sjostedti",
      ant_species == "RRB" ~ "Crematogaster mimosae",
      ant_species == "BBR" ~ "Crematogaster nigriceps"
    )
  )

# generate model predictions with standard errors
model_predictions <- predict(
  model_2,
  newdata = prediction_data,
  type = "link",
  se.fit = TRUE
)

# convert predictions from log-odds to probability scale
prediction_data <- prediction_data |>
  mutate(
    fit = plogis(model_predictions$fit),
    lower_ci = plogis(model_predictions$fit - 1.96 * model_predictions$se.fit),
    upper_ci = plogis(model_predictions$fit + 1.96 * model_predictions$se.fit)
  )

# plot observed data with model predictions and 95% confidence intervals
ggplot() +
  geom_jitter(
    data = nests_clean,
    aes(x = height_cm, y = case_control),
    width = 0,
    height = 0.03,
    alpha = 0.4,
    color = "gray40"
  ) +
  geom_ribbon(
    data = prediction_data,
    aes(x = height_cm, ymin = lower_ci, ymax = upper_ci),
    fill = "lightsalmon",
    alpha = 0.4
  ) +
  geom_line(
    data = prediction_data,
    aes(x = height_cm, y = fit),
    color = "darkred",
    linewidth = 1.2
  ) +
  facet_wrap(~ ant_species_full) +
  labs(
    x = "Tree height (cm)",
    y = "Probability of bird nest occurrence"
  ) +
  theme_classic() +
  theme(
    legend.position = "none"
  )
```

### i. 

Figure 1. Probability of bird nest occurrence increases with tree height and varies among acacia ant species. Points represent individual trees, where values of 0 indicate no bird nest was present and values of 1 indicate a bird nest was present. Dark red lines represent predicted probabilities of nest occurrence from the selected logistic regression model including tree height, acacia ant species, and their interaction. Shaded ribbons represent 95% confidence intervals around the model predictions. Panels show predictions for Crematogaster mimosae, Crematogaster nigriceps, and Crematogaster sjostedti. Data source: Lujan, E., Nielsen, R., Short, Z., Wicks, S., Nderitu Watetu, W., Malingati Khasoha, L., Palmer, T. M., Goheen, J. R., & Alston, J. M. (2023). Data, code, and metadata for: Symbiotic acacia ants drive nesting behavior by birds in an African savanna [Data set]. Zenodo. https://doi.org/10.5281/zenodo.8373322
